What it takes to create with domain-appropriate tools

### 2020-05-20 convivial-2020/acm-style/manuscript-acm.tex
Style corrections; fix random vspace; change title because QUESTION MARK >.>

\newcommand{\OROM}{Id}
\title{What does it take to create with domain-appropriate tools?}
\subtitle{Reflections on implementing the ``Id{}'' system}
\author[ ]{Joel Jakubovic}
\affiliation[ ]{University of Kent, Canterbury}

@@

Amdahl's Law generalises the following idea: even when you spend hours
of effort doubling the performance of a component used 1\% of the time,
your reward is a system overall improved by a mere 0.5\%. Now, text
coding is certainly ubiquitous, the 99\% case in programming. A small
improvement to text editing, if adopted by everyone, certainly does have
a massive \emph{intermediate} effect---but this only \emph{matters} to
the extent that text was helping us in the first place. If my goal is to
draw or animate pictures, or create a digital synth from a frequency
spectrogram, then giving me the ability to auto-indent my SVG markup is
rather underwhelming as a productivity increase, as it doesn't target
the core of the enterprise that makes it so hard.

@@

What follows in this paper is an account of how I went about building
Id, atop a preliminary ``box'' substrate.
Section\textasciitilde{}\ref{the-system} gives an overview of Id.
Section\textasciitilde{}\ref{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}
discusses the work involved in obtaining said domain-appropriate
substrate, and Section\textasciitilde{}\ref{patterns-and-polyfilling}
continues with an interpretation of the labour costs of doing this. In
Section\textasciitilde{}\ref{the-system-as-a-part-of-the-solution}, we
return to the Id system and its role in supporting flexible software.
Section\textasciitilde{}\ref{conclusions-and-future-work} addresses the
question in the title, and concludes with next steps for the project.

@@

When I first read the paper ``Open, Reusable Object Models''
\cite{OROM}, I was hooked on its idea of a small but expressive starting
system that could be self-improved into anything. It describes a
late-bound\footnote{Fewer commitments; more things determined by runtime
  conditions}, Smalltalk-style\footnote{OOP with more emphasis on object
  instances and messaging, as in a distributed system; less emphasis on
  class hierarchies implementing traditional data structures} objects
and messaging environment that the authors call ``Id''.

@@

\begin{itemize}
\tightlist
\item
  A message is sent by first \emph{bind}-ing its name to its
  \emph{implementation}: specific code, which is then run in the context
  of the receiver \(R\).
\item
  This ``bind'' step is accomplished by sending a further message; this
  time, to the receiver's \emph{vtable} \(V(R)\). A vtable is another
  object that maps ``message name'' to ``implementation code''---it's
  analogous to a ``class'' in mainstream OOP.
\item
  This initial ``bind'' message triggers a similar ``bind'' to
  \emph{its} vtable \(V(V(R))\), and so on: recursing up the
  vtable-chain, and terminating at a base case.
\item
  The higher levels of the vtable-chain mean that different kinds of
  vtables can be supported\footnote{As well as different kinds of
    ``kinds of vtables'', and so on.}, each implementing the ``bind''
  operation in its own way.
\end{itemize}

@@

For example, the first acts of the running system boil down to
initialising the three or so objects. This consists of allocating
memory, interpreting it as a C struct and then filling in fields in a
mundane manner. I had great difficulty following the specifics in my
head, but when I drew tables in the style of their diagrams I readily
saw what what going on. I personally would have preferred these to have
been in the paper in the first place, but I am not necessarily
representative of those who read it\footnote{Besides that, the
  One-Size-Fits-All{} approach is perhaps unavoidable for static print
  media.}.

At any rate, in order to solidify my understanding, there was no better
way than to run the thing and explore it. And yet, having already
``de-compiled'' English text and C source code into object diagrams, it
is a shame to have to compile it all back to struct member assignments.
Worse, the reference system does not even have text I/O when it is run,
let alone some sort of GUI\footnote{That is, the (un)intended user
  interface for Id ends up being a C debugger!}. Faced with the
necessity of adding \emph{some} UI, it seemed a waste of effort to end
up with a system that must be continually polled for its current state
at a terminal prompt.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{orom-html.png}
  \caption{Part of Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
           each other through numerical IDs. \label{fig:orom-html}}
\end{figure}

To emphasise the tendency of Id{} objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}.
Figure\textasciitilde{}\ref{fig:orom-html} shows the Id{}/HTML
implementation \cite{orom-html}, in which they take the form of HTML
tables.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{orom-svg.png}
  \caption{Part of Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
\end{figure}

In Id{}/SVG, shown in Figure\textasciitilde{}\ref{fig:orom-svg},
obj-dicts are encoded as nested SVG
\texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
of diSessa's Boxer \cite{boxer}. This was a significant departure from
the table representation, and even though SVG supports (some) nested
HTML via \texttt{\textless{}foreignObject\textgreater{}}, I actually
preferred the possibility of multiple levels of nesting.

@@

\hypertarget{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}{%
\section{Typical Requirements Of Common Software (and the work we must
do to meet
them)}\label{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../delta-transmission.png}
  \caption{Fully rigid rods transmit changes in one end to the other end. \label{fig:delta-transmission}}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../rods.png}
  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both. The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent. The remaining red rods are fully rigid. A line of CSS can reveal this ``scaffolding'' which is normally hidden for users. \label{fig:rods}}
\end{figure}

@@

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy\footnote{This highlights the mismatch between the
  tree-based DOM and any system that is graph-structured.}, but it is
still useful. However, it conflicts with my early decision to have Point
objects all share the global co-ordinate system (this was to ensure that
simple relations, such as a point following the mouse pointer, are not
infuriating to express.) Still, it was necessary in the case of certain
elements---especially those which must transcend the tree structure
altogether, like arrows between boxes---to bite this bullet, one way or
another.

@@

In SVG, we express a rectangle by
\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
The SVG specification \emph{defines} to the user that a rect simply
\textbf{is} a top-left corner, a width, and a height---and that's it.
However, this SVG-approved parametrisation of a rectangle is far from
the only one, and thus is, unsurprisingly, ill-fitted to some contexts.

I find it natural to resize boxes by dragging any of their four corners,
so I wanted this in Id{}/SVG. In this context, a ``rectangle'' is
\emph{seen as} four points: top-left, top-right, bottom-left,
bottom-right. Obviously this is not a \emph{minimal} description, since
given e.g.~the top-left and bottom-right, the other two points can be
inferred. But one way or another, to be able to drag any of them, all
four points must be present at \emph{some} level. This alternative
ontology was polyfilled in the form of a ``rect controls'' class that
can be attached manually to any SVG
\texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
are subscribed to the top-left Point\footnote{In the usual case where
  the \texttt{\textless{}rect\textgreater{}} is part of a box, the
  top-left instead controls the parent
  \texttt{\textless{}g\textgreater{}}'s \texttt{transform}---but the
  idea is the same.}; \texttt{width} and \texttt{height} are subscribed
to Rod lengths.

@@

In Id{}/SVG I did not quite want to re-map keys, but I did often want to
have things follow the mouse when dragged. In order to do this, I
reified the mouse pointer and its position, letting me write
\texttt{subscribe(point.position,\ pointer.position)}. The Point has an
\texttt{is-considering-me?} Observable wrapping
\texttt{onmouseover}/\texttt{onmouseout}, and the LMB is reified into
\texttt{left\_mouse\_button\_is\_down} to be explicit. The
aforementioned subscription is set up whenever
\texttt{is-considering-me?} and \texttt{left\_mouse\_button\_is\_down}
become true, and torn down otherwise. I tended to think of this in the
form ``subscribe to pointer \emph{only when} pointer is-considering-me
\emph{and} LMB is down'', but I could live with this notation as a
future polyfill in JavaScript.

The way these things are connected to the browser's event listeners
could be called ``device drivers''---an approach described in
\cite{prog21-bbox}. It amounts to translating information from the Web's
ontology into that of my substrate, as early as possible:

\begin{lstlisting}
svg.onmousedown = e =>
  if (e.button === 0) change(left_mouse_button_is_down, true);

@@

svg.onmousemove = e => {
  let r = svg.getBoundingClientRect();
  let pos = vsub([e.clientX, e.clientY], [r.left, r.top]);
  change(pointer.position, pos);
};
\end{lstlisting}

@@

There is, in fact, a part of the SVG specification \cite{svg-rect} that
comes \emph{frustratingly close} to the above. It spells out how the
\texttt{\textless{}rect\textgreater{}} parameters ``reduce'' to more
primitive line-drawing steps, including expressions for the four corners
in terms of \texttt{x},\texttt{y},\texttt{width},\texttt{height}{}. The
problem is, this reduction is still described in \emph{natural language}
for a human implementor, missing the potential of such a
formulation\footnote{The point here is that a machine-readable
  description could easily remain human-readable, but a natural-language
  description is not easily machine-readable.}.

@@

The desideratum of ``ontology co-existence'' is very much an open
research problem. \cite{crit-semprola} gives a taste of the thorny
complications unaddressed by a ``Semiotic Programming'' solution
\cite{semprola}, while \cite{kell-c} points out how the idea of
``linking'' is unconcerned with whatever language compiled to the object
code; language-agnosticism is one salient example of ontology
co-existence. This is especially clear if we agree with the latter
author's view that ``which language'' ought to be an
\emph{implementation detail}, that languages ought to be treatable as
\emph{different views} onto the same thing.

@@

Time and time again we come across the same pattern of partitioning
system state: trees or graphs of dictonaries, a.k.a. Maps, a.k.a.
associative arrays. I am talking about filesystem paths
\texttt{/path/to/some/file}, Python / Java modules
\texttt{com.example.pkg.subpkg}, JavaScript objects
\texttt{window.my\_obj.component}. The common case is an association of
a textual name to an arbitrary value. I find it useful to see this as a
mathematical ``function'' defined \emph{extensionally} by listing its
input/output mappings---this opposed to an \emph{intensional} definition
such as \(x \mapsto 2x+3\), or a computer program.

Extensional functions are perhaps the most basic form of Knowledge
Representation, and match natural language very well. ``The bicycle's
wheels' spokes are silver'' straightforwardly translates to a function
equation

\begin{lstlisting}
root (bicycle) (wheels) (spokes) (colour) = root (silver)
\end{lstlisting}

@@

Normally, this can be as simple as autosave to the filesystem. But the
Web platform is very wary of this\footnote{As anyone who learns WebGL
  can attest to, when they discover they must run a local Web server to
  provide image files for textures since any filesystem requests will be
  rejected for security.}, so the solution I turned to was manually
copying the markup in the browser's inspector
(Figure\textasciitilde{}\ref{fig:html-inspector}).

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\textwidth]{../html-inspector.png}
  \caption{To save the state of the system, the SVG group corresponding to the \texttt{root} box is copied and pasted inside the HTML file's \texttt{\textless{}svg\textgreater{}} element.\label{fig:html-inspector}}
\end{figure}

@@

The allure of meta-circularity here is not merely in being ``cool'', but
that it paves the way to end-user empowerment. This is somewhat
misleading because in Id{}, these ``end-users'' are actually
programmers. However, its purpose is to free a programmer's dependence
on some distant and busy\footnote{Not to mention \emph{unelected?}
  ``Take Back Control!'' (jokes aside, in this context it makes a lot of
  sense.)} language designer. It seems plausible that this is a step on
the way to enabling \emph{creation} of a piece of software which
non-programmers do not depend on us for.

@@

The Id{} paper was part of the STEPS project of Alan Kay's
VPRI\footnote{Viewpoints Research Institute}. This project argued that
the immense level of ``accidental complexity'' present in software
implementation could be reduced, and Kay himself dreams of an end result
analogous to a ``Maxwell's Equations'' of software. That is: the
behaviour of electromagnetic fields can be represented in four short
equations that fit ``on a T-shirt''.

@@

This argument as stated suffers by glossing over an important fact.
Maxwell's equations certainly do fit onto a T-shirt, but most people
will not be able to explain what they mean. What typically amounts to
years of study is compressed into those mathematical symbols, and the
learning material involved most certainly does \emph{not} fit on a
T-shirt. The obvious \emph{reductio ad absurdum} is where we encapsulate
these equations under a single symbol\footnote{In fact, this is almost
  achieved by the formalism of Geometric Calculus, which reduces them to
  only one equation.}, \(\mathcal{M}\). \(\mathcal{M}\) is defined as
``Maxwell's Equations are true''. Ta-da---this fits on a coin, but good
luck doing anything with it.

@@

Nevertheless, I wanted to explicitly highlight the obstacles one is
likely to meet when starting from this platform. They include not only
those specific to the Web, but also issues that occur far more widely
(e.g.~monopoly ontologies.) I showed how, in the short-term, we can
overcome these with solutions intended to be domain-appropriate to their
use (within the syntactic limits of JavaScript).

>>>>

\newcommand{\OROM}{Id}
\title{What it takes to create with domain-appropriate tools}
\subtitle{Reflections on implementing the ``Id{}'' system}
\author[ ]{Joel Jakubovic}
\affiliation[ ]{University of Kent, Canterbury}

@@

Amdahl's Law generalises the following idea: even when you spend hours
of effort doubling the performance of a component used 1\% of the time,
your reward is a system overall improved by a mere 0.47\%. Now, text
coding is certainly ubiquitous, the 99\% case in programming. A small
improvement to text editing, if adopted by everyone, certainly does have
a massive \emph{intermediate} effect---but this only \emph{matters} to
the extent that text was helping us in the first place. If my goal is to
draw or animate pictures, or create a digital synth from a frequency
spectrogram, then giving me the ability to auto-indent my SVG markup is
rather underwhelming as a productivity increase, as it doesn't target
the core of the enterprise that makes it so hard.

@@

What follows in this paper is an account of how I went about building
Id, atop a preliminary ``box'' substrate.
Section\textasciitilde{}\ref{the-id-system} gives an overview of Id.
Section\textasciitilde{}\ref{typical-requirements-of-common-software}
discusses the work involved in obtaining said domain-appropriate
substrate, and Section\textasciitilde{}\ref{patterns-and-polyfilling}
continues with an interpretation of the labour costs of doing this. In
Section\textasciitilde{}\ref{the-id-system-as-a-part-of-the-solution},
we return to the Id system and its role in supporting flexible software.
Section\textasciitilde{}\ref{conclusions-and-future-work} addresses the
question in the title\footnote{The original title of this paper was
  ``What does it take to create with domain-appropriate tools?'', but it
  had to be changed for technical reasons.}, and concludes with next
steps for the project.

@@

When I first read the paper ``Open, Reusable Object Models''
\cite{OROM}, I was hooked on its idea of a small but expressive starting
system that could be self-improved into anything. It describes a
late-bound,\footnote{Fewer commitments; more things determined by
  runtime conditions.} Smalltalk-style\footnote{OOP with more emphasis
  on object instances and messaging, as in a distributed system; less
  emphasis on class hierarchies implementing traditional data
  structures.} objects and messaging environment that the authors call
``Id''.

@@

\begin{itemize}
\tightlist
\item
  A message is sent by first \emph{bind}-ing its name to its
  \emph{implementation}: specific code, which is then run in the context
  of the receiver \(R\).
\item
  This ``bind'' step is accomplished by sending a further message; this
  time, to the receiver's \emph{vtable} \(V(R)\). A vtable is another
  object that maps ``message name'' to ``implementation code''---it's
  analogous to a ``class'' in mainstream OOP.
\item
  This initial ``bind'' message triggers a similar ``bind'' to
  \emph{its} vtable \(V(V(R))\), and so on: recursing up the
  vtable-chain, and terminating at a base case.
\item
  The higher levels of the vtable-chain mean that different kinds of
  vtables can be supported,\footnote{As well as different kinds of
    ``kinds of vtables'', and so on.} each implementing the ``bind''
  operation in its own way.
\end{itemize}

\vspace{-0.1em}

@@

For example, the first acts of the running system boil down to
initialising the three or so objects. This consists of allocating
memory, interpreting it as a C struct and then filling in fields in a
mundane manner. I had great difficulty following the specifics in my
head, but when I drew tables in the style of their diagrams I readily
saw what what going on. I personally would have preferred these to have
been in the paper in the first place, but I am not necessarily
representative of those who read it.\footnote{Besides that, the
  One-Size-Fits-All{} approach is perhaps unavoidable for static print
  media.}

At any rate, in order to solidify my understanding, there was no better
way than to run the thing and explore it. And yet, having already
``de-compiled'' English text and C source code into object diagrams, it
is a shame to have to compile it all back to struct member assignments.
Worse, the reference system does not even have text I/O when it is run,
let alone some sort of GUI.\footnote{That is, the (un)intended user
  interface for Id ends up being a C debugger!} Faced with the necessity
of adding \emph{some} UI, it seemed a waste of effort to end up with a
system that must be continually polled for its current state at a
terminal prompt.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{orom-html.png}
  \caption{Part of Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
           each other through numerical IDs. \label{fig:orom-html}}
\end{figure}

To emphasise the tendency of Id{} objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}.
Figure\textasciitilde{}\ref{fig:orom-html} shows the Id{}/HTML\footnote{This
  is read ``Id over HTML'', to emphasise the specific substrate on which
  it sits.} implementation \cite{orom-html}, in which they take the form
of HTML tables.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{orom-svg.png}
  \caption{Part of Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
\end{figure}

In Id{}/SVG, shown in Figure\textasciitilde{}\ref{fig:orom-svg},
obj-dicts are encoded as nested SVG
\texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
of Boxer \cite{boxer}. This was a significant departure from the table
representation, and even though SVG supports (some) nested HTML via
\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
the possibility of multiple levels of nesting.

@@

\hypertarget{typical-requirements-of-common-software}{%
\section{Typical Requirements Of Common
Software}\label{typical-requirements-of-common-software}}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{../delta-transmission.png}
  \caption{Fully rigid rods transmit changes in one end to the other end. \label{fig:delta-transmission}}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.85\textwidth]{rods.png}
  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both. The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent. The remaining red rods are fully rigid. A line of CSS can reveal this ``scaffolding'' which is normally hidden for users. \label{fig:rods}}
\end{figure*}

@@

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy,\footnote{This highlights the mismatch between the
  tree-based DOM and any system that is graph-structured.} but it is
still useful. However, it conflicts with my early decision to have Point
objects all share the global co-ordinate system (this was to ensure that
simple relations, such as a point following the mouse pointer, are not
infuriating to express). Still, it was necessary in the case of certain
elements---especially those which must transcend the tree structure
altogether, like arrows between boxes---to bite this bullet, one way or
another.

@@

In SVG, we express a rectangle by

\begin{lstlisting}
<rect x="10" y="10" width="600" height="400">
\end{lstlisting}

The SVG specification \emph{defines} to the user that a rect simply
\textbf{is} a top-left corner, a width, and a height---and that's it.
However, this SVG-approved parametrisation of a rectangle is far from
the only one, and thus is, unsurprisingly, ill-fitted to some contexts.

I find it natural to resize boxes by dragging any of their four corners,
so I wanted this in Id{}/SVG. In this context, a ``rectangle'' is
\emph{seen as} four points: top-left, top-right, bottom-left,
bottom-right. Obviously this is not a \emph{minimal} description, since
given e.g.~the top-left and bottom-right, the other two points can be
inferred. But one way or another, to be able to drag any of them, all
four points must be present at \emph{some} level. This alternative
ontology was polyfilled in the form of a ``rect controls'' class that
can be attached manually to any SVG
\texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
are subscribed to the top-left Point;\footnote{In the usual case where
  the \texttt{\textless{}rect\textgreater{}} is part of a box, the
  top-left instead controls the parent
  \texttt{\textless{}g\textgreater{}}'s \texttt{transform}---but the
  idea is the same.} \texttt{width} and \texttt{height} are subscribed
to Rod lengths.

@@

In Id{}/SVG I did not quite want to re-map keys, but I did often want to
have things follow the mouse when dragged. In order to do this, I
reified the mouse pointer and its position, letting me write
\texttt{subscribe(point.position,\ pointer.position)}. The Point has an
\texttt{is-considering-me?} Observable wrapping \texttt{onmouseover} and
\texttt{onmouseout}, while a \texttt{LMB\_is\_down} Observable reifies
the LMB state. The aforementioned subscription is set up whenever
\texttt{is-considering-me?} and \texttt{LMB\_is\_down} become true, and
torn down otherwise. I tended to think of this in the form ``subscribe
to pointer \emph{only when} pointer is-considering-me \emph{and} LMB is
down'', but I could live with this notation as a future polyfill in
JavaScript.

The way these things are connected to the browser's event listeners
could be called ``device drivers''---an approach described in
\cite{prog21-bbox}. It amounts to translating information from the Web's
ontology into that of my substrate, as early as possible
(Listing\textasciitilde{}\ref{lst:drvs}).

\begin{lstlisting}[
  label={lst:drvs},float,captionpos=b,
  caption={``Device drivers'' in the code translate the Web's ontology into a more structured form.},
]
svg.onmousedown = e =>
  if (e.button === 0)
    change(left_mouse_button_is_down, true);

@@

svg.onmousemove = e => {
  let client_pos = [e.clientX, e.clientY];
  let r = svg.getBoundingClientRect();
  let pos = vsub(client_pos, [r.left, r.top]);
  change(pointer.position, pos);
};
\end{lstlisting}

@@

There is, in fact, a part of the SVG specification \cite{svg-rect} that
comes \emph{frustratingly close} to the above. It spells out how the
\texttt{\textless{}rect\textgreater{}} parameters ``reduce'' to more
primitive line-drawing steps, including expressions for the four corners
in terms of \texttt{x},\texttt{y},\texttt{width},\texttt{height}{}. The
problem is, this reduction is still described in \emph{natural language}
for a human implementor, missing the potential in such a
formulation.\footnote{The point here is that a machine-readable
  description could easily remain human-readable, but a natural-language
  description is not easily machine-readable.}

@@

The desideratum of ``ontology co-existence'' is very much an open
research problem. \cite{crit-semprola} gives a taste of the thorny
complications unaddressed by a ``Semiotic Programming'' solution
\cite{semprola}, while \cite{kell-c} points out how the idea of
``linking'' is unconcerned with whatever language compiled to the object
code; language-agnosticism is one salient example of ontology
co-existence. This is especially clear if we agree with the latter
author's view that ``which language'' ought to be an
\emph{implementation detail}, that languages ought to be merely
\emph{different views} onto the same thing.

@@

Time and time again we come across the same pattern of partitioning
system state: trees or graphs of dictonaries, a.k.a. Maps, a.k.a.
associative arrays. I am talking about the following pattern:

\begin{itemize}
\tightlist
\item
  Filesystem paths \texttt{/path/to/some/file}
\item
  Python / Java modules \texttt{com.example.pkg.subpkg}
\item
  JavaScript objects \texttt{window.my\_obj.component}
\end{itemize}

The common case is an association of a textual name to an arbitrary
value. I find it useful to see this as a mathematical ``function''
defined \emph{extensionally} by listing its input/output mappings---this
opposed to an \emph{intensional} definition such as \(x \mapsto 2x+3\),
or a computer program.

Extensional functions are perhaps the most basic form of Knowledge
Representation, and match natural language very well. ``The bicycle's
wheels' spokes are silver'' straightforwardly translates to a function
equation: \begin{equation*}
\mathrm{root} (\mathrm{bicycle}) (\mathrm{wheels}) (\mathrm{spokes}) (\mathrm{colour}) = \mathrm{root} (\mathrm{silver})
\end{equation*}

@@

Normally, this can be as simple as autosave to the filesystem. But the
Web platform is very wary of this,\footnote{As anyone who learns WebGL
  can attest to, when they discover they must run a local Web server to
  provide image files for textures (since any filesystem requests will
  be rejected for security).} so the solution I turned to was manually
copying the markup in the browser's inspector
(Figure\textasciitilde{}\ref{fig:html-inspector}).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{../html-inspector.png}
  \caption{To save the state of the system, the SVG group corresponding to the \texttt{root} box is copied and pasted inside the HTML file's \texttt{\textless{}svg\textgreater{}} element.\label{fig:html-inspector}}
\end{figure}

@@

The allure of meta-circularity here is not merely in being ``cool'', but
that it paves the way to end-user empowerment. This is somewhat
misleading because in Id{}, these ``end-users'' are actually
programmers. However, its purpose is to free a programmer's dependence
on some distant and busy\footnote{Not to mention \emph{unelected?}
  ``Take Back Control!'' (Jokes aside, in this context it makes a lot of
  sense.)} language designer. It seems plausible that this is a step on
the way to enabling \emph{creation} of a piece of software which
non-programmers do not depend on us for.

@@

The Id{} paper was part of the STEPS project of Alan Kay's
VPRI.\footnote{Viewpoints Research Institute} This project argued that
the immense level of ``accidental complexity'' present in software
implementation could be reduced, and Kay himself dreams of an end result
analogous to a ``Maxwell's Equations'' of software. That is: the
behaviour of electromagnetic fields can be represented in four short
equations that fit ``on a T-shirt''.

@@

This argument as stated suffers by glossing over an important fact.
Maxwell's equations certainly do fit onto a T-shirt, but most people
will not be able to explain what they mean. What typically amounts to
years of study is compressed into those mathematical symbols, and the
learning material involved most certainly does \emph{not} fit on a
T-shirt. The obvious \emph{reductio ad absurdum} is where we encapsulate
these equations under a single symbol,\footnote{In fact, this is almost
  achieved by the formalism of Geometric Calculus, which reduces them to
  only one equation.} \(\mathcal{M}\). \(\mathcal{M}\) is defined as
``Maxwell's Equations are true''. Ta-da---this fits on a coin, but good
luck doing anything with it.

@@

Nevertheless, I wanted to explicitly highlight the obstacles one is
likely to meet when starting from this platform. They include not only
those specific to the Web, but also issues that occur far more widely
(e.g.~monopoly ontologies). I showed how, in the short-term, we can
overcome these with solutions intended to be domain-appropriate to their
use (within the syntactic limits of JavaScript).
